---
title: "Download and setup Seurat object"
output: html_document
---



```{r}
setwd("/home/inf-38-2022/MScProject/Main_dir")
library(tidyr)
library(dplyr)
library(Seurat)
library(ggplot2)
library(patchwork)
library(ggridges)
library(gtExtras)
source("scripts/SexAssignment.R")
```

# Download data
It is v2 of the bone marrow atlas from DISCO.
```{r}
# download from url
url <- "https://zenodo.org/record/7615517/files/disco_bone_marrow_v2.0.rds?download=1"
dir.create("data", showWarnings = FALSE)
# increase timeout to 10000 seconds
options(timeout = 10000)
# download file
# download.file(url, destfile = "data/disco_bone_marrow_v2.0.rds")
```
If it fails, manual download


# Add preprocessing steps


Load object

```{r}
data <- readRDS("data/disco_bone_marrow_v2.0.rds")
# restore data slot
data@assays$RNA@data <- data@assays$RNA@counts
```


Add variable features, scale data and run PCA
```{r}
data <- FindVariableFeatures(data, selection.method = "vst", nfeatures = 2000)
data <- ScaleData(data, verbose = FALSE ) # scales data only for variable features
data <- RunPCA(data, verbose = FALSE)
```

## Add extra metadata
Group cell types into lineages.
```{r}
# lineages
lymphoid_lineage <- c("Naive B cell", "Memory B cell", "Plasma cell", "preB cell", "preB cell (cycling)", "proB cell",
                      "proB cell (cycling)", "Plasma cell", "Naive CD8 T cell", "Effector/Memory CD4 T cell",
                      "Naive CD4 T cell", "GZMK+ CD8 T cell", "Gamma-delta T cell", "IFN-activated T cell", "Treg",
                      "GZMB+ CD8 T cell", "MAIT", "T/NK cell (cycling)", "CD16+ NK", "Tissue-resident NK cell",
                      "CD56+ NK", "ILC", "CLP")
gmp_lineage <- c("GMP", "CD14+MHCIIhigh monocyte", "CD14+MHCIIlow monocyte", "Macrophage", "CD16+ monocyte",
                 "S100A+ preNeutrophil (cycling)", "S100A+ preNeutrophil", "Metamyelocyte/Band neutrophil",
                 "Neutrophil", "Promyelocyte", "Myelocyte", "Mast cell")
dendritic_cell_lineage <- c("pre-pDC (myeloid origin)", "cDC2", "cDC2 (cycling)", "cDC1", "pDC", "CDP",
                            "pre-pDC (lymphoid origin)" )
mep_lineage <- c("MEP", "Early SOX4+ erythroblast", "Late hemoglobin+ erythroblast", "Fetal HBG+ erythrocyte",
                 "Intermediate EPCAM+ erythroblast", "Erythrocyte", "Megakaryocyte", "Megakaryocyte progenitor",
                 "Platelet")
early_progenitor_cells <- c("CMP", "HSC", "MPP")
stroma <- c("Fibroblast", "Mesenchymal stromal cell", "Osteoclast")
lineages <- ifelse(data$ct %in% lymphoid_lineage, "Lymphoid",
                   ifelse(data$ct %in% gmp_lineage, "GMP",
                          ifelse(data$ct %in% dendritic_cell_lineage, "Dendritic cell",
                                 ifelse(data$ct %in% mep_lineage, "MEP",
                                        ifelse(data$ct %in% early_progenitor_cells, "Early progenitor cells",
                                               ifelse(data$ct %in% stroma, "Stroma", "Unknown"))))))
# add to seurat object
data <- AddMetaData(data, metadata = lineages, col.name = "lineage")

```

Group cell types into simplified cell types
```{r}
# simplified_ct
b_cells <- c("Naive B cell", "Memory B cell", "Plasma cell", "preB cell", "preB cell (cycling)", "proB cell",
             "proB cell (cycling)", "Plasma cell")
t_nk_cells<- c("Naive CD8 T cell", "Effector/Memory CD4 T cell", "Naive CD4 T cell",
               "GZMK+ CD8 T cell", "Gamma-delta T cell", "IFN-activated T cell", "Treg", "GZMB+ CD8 T cell", "MAIT",
               "T/NK cell (cycling)","CD16+ NK", "Tissue-resident NK cell", "CD56+ NK","ILC")
erythrocytes <- c("Early SOX4+ erythroblast", "Late hemoglobin+ erythroblast", "Fetal HBG+ erythrocyte",
                  "Intermediate EPCAM+ erythroblast", "Erythrocyte")
monocytes <- c("CD14+MHCIIhigh monocyte", "CD14+MHCIIlow monocyte", "Macrophage", "CD16+ monocyte")
granulocyte <- c("S100A+ preNeutrophil (cycling)", "S100A+ preNeutrophil", "Metamyelocyte/Band neutrophil",
                 "Neutrophil", "Promyelocyte", "Myelocyte", "Mast cell")
dendritic_cells <- c("pre-pDC (lymphoid origin)", "pre-pDC (myeloid origin)", "cDC2", "cDC2 (cycling)", "cDC1", "pDC")
Mk <- c("Megakaryocyte", "Megakaryocyte progenitor", "Platelet")
progenitor_cells <- c("CLP", "CMP", "GMP", "MEP", "MPP", "CDP", "HSC")
stroma <- c("Fibroblast", "Mesenchymal stromal cell", "Osteoclast")

# add a column with the simplified cell type name
simplified_ct <- ifelse(data$ct %in% b_cells, "B cells",
                        ifelse(data$ct %in% t_nk_cells, "T/NK cells",
                               ifelse(data$ct %in% erythrocytes, "Erythrocytes",
                                      ifelse(data$ct %in% monocytes, "Monocytes",
                                             ifelse(data$ct %in% granulocyte, "Granulocytes",
                                                    ifelse(data$ct %in% dendritic_cells, "Dendritic cells",
                                                           ifelse(data$ct %in% Mk, "Mk",
                                                                  ifelse(data$ct %in% progenitor_cells, "Progenitor cells",
                                                                         ifelse(data$ct %in% stroma, "Stroma",
                                                                                "Unknown")))))))))
# add to seurat object
data <- AddMetaData(data, metadata = simplified_ct, col.name = "simplified_ct")

```


# Predict sex

First run SexAssign, a function that is able to assign sex to samples
```{r}
# Run SexAssign
sex_assign_out <- SexAssign(data = data,
                            sex_col = "gender", # previous sex annotation
                            sample_col = "sample", # sample column
                            do.report = TRUE,
                            assay = "RNA",
                            min.percent = 0.7, # minimum percentage of cells to be considered of one sex
                            min.ratio = 2 # mininum ratio between the sexes for it to be considered of the sex in the numerator
)
```
If some "features" can't be found, it means that these genes aren't in the gene expression matrix.

Explore the output
```{r}
# Table per sample
sex_assign_out[[2]]
# Focus on the inconclusive samples
sex_assign_out[[2]] %>%
        filter(sample_sex == "inconclusive")
# Frequency plot
sex_assign_out[[3]] -> freq_SexAssign_plot
freq_SexAssign_plot
# sex genes heatmap
sex_assign_out[[4]] -> heatmap_SexAssign_plot
heatmap_SexAssign_plot
```


Save the new object as data
```{r}
data <- sex_assign_out[[1]]
save <- sex_assign_out[-1]
saveRDS(save, "results/output_obj/sex_assign_out.rds")
rm(save)
```

Compare sex annotation with prediction.
```{r}
prev_sex <- data@meta.data %>%
        distinct(sample, gender) %>%
        dplyr::count(gender) %>%
        mutate(var = "Provided\nsex annotation") %>%
        dplyr::rename(Predicted_sex = gender) %>%
        mutate( Predicted_sex = ifelse(Predicted_sex == "", "Not provided", Predicted_sex))
Pred_sex <- data@meta.data %>%
        distinct(sample, Predicted_sex) %>%
        dplyr::count(Predicted_sex) %>%
        mutate(var="SexAssign\nprediction")
sex_annot <- bind_rows(prev_sex, Pred_sex)
sex_annot_bar <- sex_annot %>%
        ggplot( aes(fill=factor(Predicted_sex, c("F", "Not provided", "inconclusive", "M")), y=n, x=var, label = n)) +
        geom_bar(position="fill", stat="identity") +
        geom_text(size = 8, position = position_fill(vjust = 0.5)) +
        # change text
        labs(x = "",
             y = "Proportion of samples",
             # title = "Sex annotation vs. SexAssign prediction",
             fill = "Sample sex") +
        theme_classic() +
        scale_fill_manual(values = c("M" = "gold", "F" = "purple", "Not provided" = "darkgray", "inconclusive" = "lightgray"),
                          labels = c( "inconclusive" = "Inconclusive", "F" = "Female", "M" = "Male"))
        # change text size
        # theme(text = element_text(size=20),
        #       axis.text.x = element_text(size = 15),
        #       axis.text.y = element_text(size = 20),
        #       legend.text = element_text(size = 20),
        #       legend.title = element_text(size = 20),
        #       plot.title = element_text(size = 20, face = "bold"))

mismatch_barplot <- as.data.frame(sex_assign_out[[2]]) %>%
  # make match column factor
    mutate(match = factor(match, c("New_annotation", "Mismatch",  "Match" ))) %>%
  dplyr::count(match) %>%
  mutate(var="") %>%
  ggplot(aes(fill=factor(match, c("Match", "Mismatch", "New_annotation")), x=var, y=n, label = n)) +
    geom_bar(position="fill", stat="identity") +
    geom_text(size = 8, position = position_fill(vjust = 0.5)) +
    # change text
    labs(x="",
         y="Proportion of samples",
         # title = "Mismatches in sample sex",
         fill = "") +
    theme_classic() +
    scale_fill_manual(values = c("Match" = "#90ee90", "Mismatch" = "#ff0033", "New_annotation" = "#4169E1"),
                      labels = c("New_annotation" = "New annotation"))

        # theme(text = element_text(size=20),
        #       axis.text.x = element_text(size = 20),
        #       axis.text.y = element_text(size = 20),
        #       legend.text = element_text(size = 20),
        #       legend.title = element_text(size = 20),
        #       plot.title = element_text(size = 20, face = "bold"))



sex_annot_bar + mismatch_barplot
```

# Plot the data
## ggextra
```{r}
gt_plt_summary(data@meta.data)

```




## UMAP
```{r}
# create base plot
plot1 <- DimPlot(data, group.by ="ct", raster = FALSE) + NoLegend()
plot1
# add labels in color
plot2 <- LabelClusters(plot1, id="ct", color="black", size = 5, repel=T, max.overlaps=Inf) # ident == ct
plot2
# add labels in black
plot3 <- LabelClusters(plot1, id="ct", color="ct", size = 5, repel=T, max.overlaps=Inf)
plot3
```


## age distribution

Create summary table
```{r}
# number of samples per project
sample_per_proj <- data@meta.data %>%
        distinct(project_id, sample, age) %>%
        group_by(project_id) %>%
        summarise(n_samples = n(),
                  mean_age = mean(age),
                  sd_age = sd(age),
                  min_age = min(age),
                  max_age = max(age),)
sample_per_proj
```

### age distribution plots
```{r}
# per project
age_distr_proj <- data@meta.data %>%
        ggplot(aes(x = age, y=..density.., fill = factor(project_id))) +
        geom_density(alpha = 0.5) +
        labs(x = "Age", y = "Density",
             title = "Age distribution for every project in DISCO Bone Marrow Atlas v2")
age_distr_proj

# per Predicted_sex
age_distr_sex <- data@meta.data %>%
        ggplot(aes(x = age, y=..density.., fill = factor(Predicted_sex))) +
        geom_density(alpha = 0.5) +
        labs(x = "Age", y = "Density",
             title = "Age distribution for Males and Females in DISCO Bone Marrow Atlas v2")
age_distr_sex

# per Predicted_sex and faceted by project
age_distr_sex_proj <- data@meta.data %>%
        ggplot(aes(x = age, y=..density.., fill = factor(Predicted_sex))) +
        geom_density(alpha = 0.5) +
        facet_wrap(~project_id) +
        labs(x = "Age", y = "Density",
             title = "Age distribution for Males and Females in DISCO Bone Marrow Atlas v2")
age_distr_sex_proj
age_distr_sex_proj <- data@meta.data %>%
        ggplot(aes(x = age, y=project_id, fill = factor(Predicted_sex))) +
        geom_density_ridges(alpha = 0.5, rel_min_height = 0.01) +
        labs(x = "Age", y = "Density",
             title = "Age distribution for Males and Females in DISCO Bone Marrow Atlas v2 per project",
             fill = "Sex")
age_distr_sex_proj

```

# Patchwork make figures

```{r}
fig2_patchwork <- (heatmap_SexAssign_plot  + mismatch_barplot )/ freq_SexAssign_plot
fig2_patchwork + plot_annotation(
        title = "Sex assignment results for DISCO Bone Marrow Atlas v2",
        subtitle = "Done with SexAssign based on cellXY",
        caption = "some caption"
)
```

```{r}


supl_fig_heatmap_sex <- sex_assign_out[[4]] +
labs(
     caption = "Suplementary Figure 1. Gene expression corroborates SexAssign+cellXY sex prediction. Genes are the ones used by cellXY models. Missing genes are not present in the gene expression matrix. Samples have been ordered by SexAssign prediction (Female, Male, and Inconclusive).") +
               theme(plot.title = element_text(size = 15),
                     plot.margin = margin(l=10, r = 20),
                     plot.caption = ggtext::element_textbox_simple(
                       size = 10, hjust = 0, margin = margin(t = 3, r = 0, b = 0, l = 0))) +
  plot_annotation(title = "Heatmap of mean expression of X and Y genes per sample in DISCO Bone Marrow Atlas v2.")
supl_fig_heatmap_sex
ggsave("results/plots/01_supl_01_heatmap.png", supl_fig_heatmap_sex, dpi = 600, height = 9, width = 12)

```

figure 1 about mismatches
```{r}
fig1 <- patchwork::wrap_elements(sex_annot_bar) + patchwork::wrap_elements(mismatch_barplot) +
  plot_annotation( tag_levels = "A",
  caption = "Figure 2. SexAssign effectively annotates sex and reveals dubious sex annotation in the DISCO Bone Marrow Atlas v2. A. Sex sample metadata from the provided sex annotation and the sex prediction by SexAssign. B. Overlap of the two annotations.") &
  theme( plot.caption = ggtext::element_textbox_simple(size = 10), plot.title =ggtext::element_textbox_simple(size = 1))
fig1

ggsave("results/plots/01_fig_02_mismatches.png", fig1, dpi = 600, height = 6, width = 13)
```


# save
The object
```{r}
dir.create("data/processed", showWarnings = FALSE)
saveRDS(data, file = "data/processed/DISCO_BM_v2_processed.rds", compress = FALSE)
```

Plots and tables

```{r}
dir.create("results/", showWarnings = FALSE) # create results folder
dir.create("results/plots", showWarnings = FALSE)
dir.create("results/tables", showWarnings = FALSE)


ggsave("results/plots/01_sex_freq.png", sex_assign_out[[3]], dpi=600)
ggsave("results/plots/01_sex_heatmap.png", sex_assign_out[[4]], dpi=600)
ggsave("results/plots/01_umap_ct.png", plot2, dpi=600)
ggsave("results/plots/01_umap_ct_black.png", plot3, dpi=600)
ggsave("results/plots/01_age_distr_proj.png", age_distr_proj, dpi=600)
ggsave("results/plots/01_age_distr_sex.png", age_distr_sex, dpi=600)
ggsave("results/plots/01_age_distr_sex_proj.png", age_distr_sex_proj, dpi=600)
ggsave("results/plots/01_mismatch_barplot.png", width = 10, height = 6,mismatch_barplot, dpi=600)

write.table(sex_assign_out[[2]] , file = "results/tables/01_sample_sex_res.csv", sep = ",", quote = FALSE, row.names = FALSE)
write.table(sample_per_proj, file = "results/tables/01_sample_per_proj.csv", sep = ",", quote = FALSE, row.names = FALSE)
```
```{r}
# save the session info
sessionInfo()
# to file
sink("results/tables/01_session_info.txt")
```
