---
title: "R Notebook"
output: pdf_document
---
```{r include = FALSE}
knitr::opts_chunk$set(fig.height = 9, fig.width = 7, fig.crop = FALSE)
```
This is based on the tutorial
https://github.com/IanevskiAleksandr/sc-type

# load libraries and functions

```{r load libraries and functions, echo=FALSE}
rm(list=ls())
suppressMessages(library(dplyr))
suppressMessages(library(Seurat))
suppressMessages(library(HGNChelper))
suppressMessages(library(openxlsx))
suppressMessages(library(dittoSeq))

source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/gene_sets_prepare.R")
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/sctype_score_.R")

```


# load data

```{r load data}
# load data
bm <- readRDS("data/seurat_obj/Oejten_seurat_QC.rds")
# load gene database, provided
gs_list <- gene_sets_prepare("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/ScTypeDB_short.xlsx", "Immune system")
```

# run sc-type

```{r run sc-type, echo=FALSE}
# run sc-type
es.max <- sctype_score(scRNAseqData = bm[["RNA"]]@scale.data,
                       scaled = TRUE,
                       gs = gs_list$gs_positive,
                       gs2 = gs_list$gs_negative)
```


# merge by cluster

```{r merge by cluster, echo=FALSE}

cL_resutls <- do.call("rbind", lapply(unique(bm@meta.data$seurat_clusters), function(cl){
    es.max.cl <- sort(rowSums(es.max[ ,rownames(bm@meta.data[bm@meta.data$seurat_clusters==cl, ])]), decreasing = !0)
    head(data.frame(cluster = cl, type = names(es.max.cl), scores = es.max.cl, ncells = sum(bm@meta.data$seurat_clusters==cl)), 10)
}))
# save results to csv
write.csv(cL_resutls, "results/tables/cL_resutls.csv")
# get top score for each cluster
sctype_scores <- cL_resutls %>% group_by(cluster) %>% top_n(n = 1, wt = scores)

# set low-confident (low ScType score) clusters to "unknown"
sctype_scores$type[as.numeric(as.character(sctype_scores$scores)) < sctype_scores$ncells/2] <- "Unknown"
print(sctype_scores[,1:4])
# save results to csv and xlsx
write.csv(sctype_scores, "results/tables/sctype_scores.csv")
write.xlsx(sctype_scores, "results/tables/sctype_scores.xlsx")
```

# plot on umap

```{r plot on umap, echo=FALSE}
bm@meta.data$sctypeclassif <- ""
for(j in unique(sctype_scores$cluster)){
  cl_type <- sctype_scores[sctype_scores$cluster==j,]
  bm@meta.data$sctypeclassif[bm@meta.data$seurat_clusters == j] <- as.character(cl_type$type[1])
}

bm %>%
    DimPlot(reduction = "umap",
            label = TRUE,
            repel = TRUE,
            group.by = 'sctypeclassif') +
            # change legend text size and put at the bottom
            theme(legend.text = element_text(size = 10),
                  legend.position = "bottom"
            ) +
            # change title
            ggtitle("sc-type") -> umap_celltype_plot
umap_celltype_plot
# save plot to png
ggsave("results/figures/umap_celltype_plot.png", umap_celltype_plot, width = 10, height = 10)
```

# plot cell type stats
```{r plot cell type stats, echo=FALSE}
# plot cell type frequency in seurat object metadata
# by sample_id
bm %>%
  dittoBarPlot(var = "sctypeclassif", group.by = "sample") +
   theme(legend.text = element_text(size = 10),
         legend.position = "bottom"
   )-> celltype_freq_plot
celltype_freq_plot
# save plot to png
ggsave("results/figures/celltype_freq_plot.png", celltype_freq_plot, width = 10, height = 10)

```


# group into lineages
```{r group into lineages, echo=FALSE}
# get unique cell types
bm@meta.data %>%
  dplyr::select(sctypeclassif) %>%
  unlist() %>%
    unique() -> cell_types
# group into lineages
bm[["lineage"]] <- ""
# create lists
myeloid_lineage <- c("Myeloid Dendritic cells", "Neutrophils", "Classical Monocytes", "Non-classical monocytes", "Basophils", "Neutrophils")
lymphoid_lineage <- c("Plasma B cells", "Immature B cells", "Pro-B cells", "Pre-B cells", "Naive CD4+ T cells", "Naive CD8+ T cells", "CD8+ NKT-like cells", "Plasmacytoid Dendritic cells",
                      "Memory CD4+ T cells", "Memory CD8+ T cells", "Naive B cells")
erythroid_lineage <- c("Erythroid-like and erythroid precursor cells", "Endothelial", "Platelets")

# Create a vector of lineage assignments for each cell based on its cell type
lineage_assignments <- ifelse(bm$sctypeclassif %in% myeloid_lineage, "Myeloid",
                              ifelse(bm$sctypeclassif %in% lymphoid_lineage, "Lymphoid",
                                     ifelse(bm$sctypeclassif %in% erythroid_lineage, "Erythroid",
                                            ifelse(bm$sctypeclassif %in% "Progenitor cells", "Progenitor cells",
                                                   "Unknown"))))
# Add the lineage metadata column to the Seurat object
bm <- AddMetaData(bm, metadata = lineage_assignments, col.name = "lineage")


# plot lineage frequency in seurat object metadata
# by sample_id
bm %>%
  dittoBarPlot(var = "lineage", group.by = "sample") -> lineage_freq_plot
lineage_freq_plot
# save plot to png
ggsave("results/figures/lineage_freq_plot.png", lineage_freq_plot, width = 10, height = 10)
```

Samples G and P have a lot more erythroid lineage cells than the other samples proportionally.
## individual clusters
```{r individual clusters, echo=FALSE}
# plot seurat_cluster frequency in seurat object metadata
# by sample
bm %>%
  dittoBarPlot(var = "seurat_clusters", group.by = "sample") -> lineage_freq_plot
lineage_freq_plot
# save plot to png
ggsave("results/figures/seurat_clusters_by_sample_frequency.png", lineage_freq_plot, width = 10, height = 10)
```
It is P again with cluster 14 which is an erythroid cluster mapping to the lymphoid lineage.

# plot lineages on umap
```{r plot lineages on umap, echo=FALSE}
bm %>%
    DimPlot(reduction = "umap",
            label = TRUE,
            repel = TRUE,
            group.by = 'lineage') +
            theme(legend.text = element_text(size = 10),
                  legend.position = "bottom"
            )-> umap_lineage_plot
umap_lineage_plot
# save plot to png
ggsave("results/figures/umap_lineage_plot.png", umap_lineage_plot, width = 10, height = 10)
```

A small subset of erythroid clusters with the lymphoid cluster in the middle.

# lineage frequency by sex
```{r}
bm %>%
  dittoBarPlot(var = "lineage", group.by = "sex") -> sexlineage_freq_plot
sexlineage_freq_plot
bm %>%
  dittoBarPlot(var = "sctypeclassif", group.by = "sex") -> sexcell_freq_plot
sexcell_freq_plot
# save plot to png
ggsave("results/figures/sexlineage_freq_plot.png", sexlineage_freq_plot, width = 10, height = 10)
```

We can see the usual CD8+ dominance in males

## compare umap by sex
```{r compare umap by sex}
bm %>%
    DimPlot(reduction = "umap",
            label = TRUE,
            repel = TRUE,
            group.by = 'lineage',
            split.by = 'sex') +
    theme(legend.text = element_text(size = 10),
          legend.position = "bottom"
    )-> sex_split_umap
sex_split_umap
# umap with just sex
bm %>%
        DimPlot(reduction = "umap",
                label=TRUE,
                repel=TRUE,
                group.by= "sex"
        ) +
        theme(legend.text = element_text(size = 10),
              legend.position = "bottom"
        )-> sex_umap
sex_umap
```


# save seurat object
```{r save seurat object, echo=FALSE}
# save seurat object
saveRDS(bm, file = "data/seurat_obj/Oejten_annotated.rds", compress = FALSE)
print("seurat object saved")
# print time
print(Sys.time())
```

# load seurat object
```{r load seurat object, echo=FALSE}
# load seurat object
bm <- readRDS("data/seurat_obj/Oejten_annotated.rds")
print("seurat object loaded")
# print time
print(Sys.time())
```


# plot cell type frequency by sample grouped by sex
```{r}
bm %>%
  dittoBarPlot(var = "sctypeclassif", group.by = "sample", split.by = "sex" ) -> sexcell_freq_plot

sexcell_freq_plot
```
