---
title: "Cell trajectory analysis using Monocle 3"
output: pdf_document
---

This follows
https://ucdavis-bioinformatics-training.github.io/2021-August-Advanced-Topics-in-Single-Cell-RNA-Seq-Trajectory-and-Velocity/data_analysis/scrna_analysis_prepare1_fixed
https://ucdavis-bioinformatics-training.github.io/2021-August-Advanced-Topics-in-Single-Cell-RNA-Seq-Trajectory-and-Velocity/data_analysis/monocle_fixed
https://cole-trapnell-lab.github.io/monocle3/docs/trajectories/

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.height = 9, fig.width = 7, fig.crop = FALSE)
```

# Loading
## load packages

```{r load libraries}
rm(list=ls())
library(monocle3)
library(Seurat)
library(SeuratWrappers)
library(patchwork)

set.seed(1234)
```
## Functions

```{r functions}
# From the tutorial cole-trapnell-lab.github.io/monocle3/docs/trajectories/
# a helper function to identify the root principal points:
get_earliest_principal_node <- function(monocle3_object, time_bin="130-170"){
  #
  cell_ids <- which(colData(monocle3_object)[, "embryo.time.bin"] == time_bin)

  closest_vertex <-
  cds@principal_graph_aux[["UMAP"]]$pr_graph_cell_proj_closest_vertex
  closest_vertex <- as.matrix(closest_vertex[colnames(monocle3_object), ])
  root_pr_nodes <-
  igraph::V(principal_graph(monocle3_object)[["UMAP"]])$name[as.numeric(names
  (which.max(table(closest_vertex[cell_ids,]))))]

  root_pr_nodes
}
# source from github v1.0.0, but modified rbind
source("data/functions/graph_test.R")

```
## load data

```{r load data}
# load data
data <- readRDS("data/seurat_obj/Oejten_annotated.rds")
```

# Seurat to monocle3

```{r Seurat to monocle3}
# convert Seurat object to monocle3 object
monocle3_obj <- as.cell_data_set(data)
```

# Cluster analysis

```{r cluster analysis}
# cluster analysis
monocle3_obj <- cluster_cells(monocle3_obj, resolution = 1e-5)
print("cluster_cells done")

```

1e-5 looks pretty good
## plot the clusters
```{r plot clusters}
p1 <- plot_cells(monocle3_obj, color_cells_by = "cluster", show_trajectory_graph = FALSE)
p1
# plot partition
plot_cells(monocle3_obj, color_cells_by = "partition", show_trajectory_graph = FALSE) -> p1.2
p1.2

```

# Pseudotime analysis

```{r pseudotime analysis}
# pseudotime analysis
monocle3_obj <- learn_graph(monocle3_obj, use_partition = TRUE)
```

## plot the trajectory
```{r plot trajectory}
plot_cells(monocle3_obj,
           color_cells_by = "cluster",
           label_groups_by_cluster=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE) -> p2
p2

```

## plot the pseudotime
```{r plot pseudotime}
# root cell is cluster 4 where we know the stem cells are
monocle3_obj <- order_cells(monocle3_obj, root_cell = colnames(monocle3_obj[,clusters(monocle3_obj) == 4]))
plot_cells(monocle3_obj,
           color_cells_by = "pseudotime",
           label_groups_by_cluster=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE) -> p3
p3
plot_cells(monocle3_obj,
           color_cells_by = "sctypeqin",
           label_groups_by_cluster=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE) -> p3.2
p3.2
```
# pick the root

## find the earliest node and order the cells from it
```{r pick the root}
# pick the root

# order cells from the root
monocle3_obj <- order_cells(monocle3_obj, root_pr_nodes=get_earliest_principal_node(monocle3_obj))

```
 ## plot the pseudotime
```{r plot pseudotime}
plot_cells(monocle3_obj,
           color_cells_by = "pseudotime",
           label_groups_by_cluster=FALSE,
           label_leaves=TRUE,
           label_branch_points=FALSE,
           graph_label_size = 4) -> p3.3
p3.3
```



# subset trajectory

```{r subset trajectory}
# subset trajectory
# the interactive selector doesn't work in pycharm (at least), so we need to do it manually
# monocle3_obj <- choose_graph_segments(monocle3_obj)

# take trajectory 9
trajectory_9 <- subset(monocle3_obj, cells = which(colData(monocle3_obj)$trajectory == 9))
# plot the pseudotime
plot_cells(trajectory_9,
           color_cells_by = "pseudotime",
           group_cells_by = "sex",
           label_groups_by_cluster=FALSE,
           label_leaves=TRUE,
           label_branch_points=FALSE,
           graph_label_size = 4
           ) -> p3.4
p3.4
```


# 3d trajectory

```{r 3d trajectory}
# 3d trajectory
cds_3d <- reduce_dimension(monocle3_obj, max_components = 3)
cds_3d <- cluster_cells(cds_3d)
cds_3d <- learn_graph(cds_3d)
cds_3d <- order_cells(cds_3d, root_pr_nodes=get_earliest_principal_node(cds))
```
## plot the 3d trajectory
```{r plot 3d trajectory}
cds_3d_plot_partition <- plot_cells_3d(cds_3d, color_cells_by="partition")
cds_3d_plot_partition
# color by sctypeqin
cds_3d_plot_qin <- plot_cells_3d(cds_3d, color_cells_by="sctypeqin")
cds_3d_plot_qin
```

# save the object
```{r save the object}
# if folder doesn't exist, create it
dir.create("data/monocle3_obj", showWarnings = FALSE)
# save the object
saveRDS(monocle3_obj, "data/monocle3_obj/Oejten_annotated.rds", compress = FALSE)
```
# load the object
```{r load the object}
# load the object
monocle3_obj <- readRDS("data/monocle3_obj/Oejten_annotated.rds")
```
# Differential expression
We can identify the genes that are differentially expressed between two groups of cells. We can also identify the genes that are differentially expressed between two groups of cells along the trajectory. We can also identify the genes that are differentially expressed between two groups of cells along the trajectory, but only in the cells that are in the same cluster as the root cell.

```{r differential expression eval = FALSE}
pr_test <- graph_test(monocle3_obj, neighbor_graph = "principal_graph", cores = 4)
pr_deg_ids <- row.names(subset(pr_test, q_value < 0.05))
pr_deg_ids
# save both to file
write.table(pr_deg_ids, file = "results/tables/pr_deg_ids.txt", sep = "\t", quote = FALSE, row.names = FALSE)
write.table(pr_test, file = "results/tables/pr_test.txt", sep = "\t", quote = FALSE, row.names = TRUE)

```
if rbind error, do this
https://groups.google.com/g/monocle-3-users/c/tBjYuAxwyEo?pli=1
The problem is that for this version of Monocle3, 1.0.0, the graph_test function uses Matrix::rBind() instead of rbind().
It has to be manually modified via trace().
The editor doesn´t work in PyCharm, but it can be done in the terminal, with the condition that this chunk has to be also run there.

### load the gene tables
```{r load the gene tables}
# load the gene tables
pr_deg_ids <- read.table("results/tables/pr_deg_ids.txt", sep = "\t", header = FALSE, stringsAsFactors = FALSE)
pr_test <- read.table("results/tables/pr_test.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE, row.names = 1)
# sort by morans_I (morans_index)
pr_deg_ids <- row.names(subset(pr_test[order(pr_test$morans_I),], q_value < 0.05))
# print the first 10
pr_deg_ids[1:12] -> markers_list
```

## plot some interesting genes


```{r plot some interesting genes}
# fix gene names
rowData(monocle3_obj)$gene_name <- rownames(monocle3_obj)
rowData(monocle3_obj)$gene_short_name <- rowData(monocle3_obj)$gene_name
# plot some interesting genes
# show only partition 2
integrated.sub_2 <- subset(as.Seurat(monocle3_obj, assay = NULL), monocle3_partitions == 2)
partition_2_cds <- as.cell_data_set(integrated.sub_2)
# fix gene names again
rowData(partition_2_cds)$gene_name <- rownames(partition_2_cds)
rowData(partition_2_cds)$gene_short_name <- rowData(partition_2_cds)$gene_name

plot_cells(partition_2_cds,
           genes= markers_list,
           show_trajectory_graph=FALSE,
           labels_per_group = FALSE,
           label_leaves=FALSE,
           min_expr = 1e-10,
           scale_to_range = TRUE
            ) -> p5
p5
# plot dot plot expression of the genes
plot_genes_by_group(monocle3_obj,
                    markers = markers_list,
                    axis_order = "marker_group",
                    group_cells_by="sctypeqin",
                    ordering_type="maximal_on_diag",
                    max.size=10) -> p5.2
p5.2
```

## plot marker genes in dot plot
```{r plot marker genes in dot plot}
markers_list <- c("HBB", "HBA2", "HBA1", "IGLV3-27", "AHSP", "PRDX2", "HBD", "LYZ", "BLVRB", "ALAS2", "MZB1", "CD3D", "CD14")
plot_genes_by_group(monocle3_obj,
                    markers = markers_list,
                    axis_order = "marker_group",
                    group_cells_by="sctypeqin",
                    ordering_type="maximal_on_diag",
                    max.size=10) -> p6
p6
```

## plot expression in pseudotime
```{r plot expression in pseudotime}
#plot_genes_in_pseudotime(monocle3_obj[markers_list,], # only the markers
#                         panel_order = markers_list,
#                         color_cells_by = "sctypeqin",
#                         min_expr = 0.1
#) -> p7
p7
plot_genes_in_pseudotime(monocle3_obj["HBB",], # only the markers
                         panel_order = "HBB",
                         color_cells_by = "sctypeqin",
                         min_expr = 0.1
) -> p7.2
p7.2
plot_genes_in_pseudotime(monocle3_obj["LYZ",], # only the markers
                         panel_order = "LYZ",
                         color_cells_by = "sctypeqin",
                         min_expr = 0.1
) -> p7.3
p7.3
# save to pdf
ggsave("results/figures/monocle3_pseudotime.pdf", p7, width = 10, height = 10)
```

### see expression
```{r see expression}
exprs(monocle3_obj)[markers_list,] -> exprs_markers
rowMeans(exprs_markers) -> exprs_markers_mean
exprs_markers_mean
plot_genes_violin(monocle3_obj["HBB",], group_cells_by = "sctypeqin", normalize = FALSE, log_scale = FALSE ) -> p8
p8
plot_genes_violin(monocle3_obj["LYZ",], group_cells_by = "sctypeqin", normalize = FALSE, log_scale = FALSE ) -> p8.2
p8.2
```



