---
title: "Cell frequency analysis"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# load
```{r load libraries}
library(dplyr)
library(ggplot2)
library(ggrepel)
library(Seurat)
library(patchwork)
library(dittoSeq)
library(speckle)
setwd("/home/inf-38-2022/MScProject/02_CellFrequency/")
source("../b00_SexAssignment/b00_SexAssignment.R")
```
load data
```{r load data}
rna.data <- readRDS("../01_Integration/rna_annotated_adult_progenitors.rds")
```

Add sex annotation
```{r}
sex_assign_out <- SexAssign(data = rna.data, sex_col = "gender", sample_col = "sample", report = TRUE,
                            assay = "raw" # use raw slot for this integrated dataset, because the RNA slot only contains variable features.
)
# see table
sex_assign_out[[2]]
# see plot
sex_assign_out[[3]]
```
Now we change the seurat object
```{r}

# add to object
rna.data <- sex_assign_out[[1]]
```
We add simplified cell types
```{r}
b_cells <- c("Naive B cell", "Memory B cell", "Plasma cell", "preB cell", "preB cell (cycling)", "proB cell", "proB cell (cycling)", "Plasma cell")
t_nk_cells <- c("Naive CD8 T cell", "Effector/Memory CD4 T cell", "Naive CD4 T cell",
                "GZMK+ CD8 T cell", "Gamma-delta T cell", "IFN-activated T cell", "Treg", "GZMB+ CD8 T cell", "MAIT", "T/NK cell (cycling)", "CD16+ NK", "Tissue-resident NK cell", "CD56+ NK", "ILC")
erythrocytes <- c("Early SOX4+ erythroblast", "Late hemoglobin+ erythroblast", "Fetal HBG+ erythrocyte",
                  "Intermediate EPCAM+ erythroblast", "Erythrocyte")
monocytes <- c("CD14+MHCIIhigh monocyte", "CD14+MHCIIlow monocyte", "Macrophage", "CD16+ monocyte")
granulocyte <- c("S100A+ preNeutrophil (cycling)", "S100A+ preNeutrophil", "Metamyelocyte/Band neutrophil",
                 "Neutrophil", "Promyelocyte", "Myelocyte", "Mast cell")
dendritic_cells <- c("pre-pDC (lymphoid origin)", "pre-pDC (myeloid origin)", "cDC2", "cDC2 (cycling)", "cDC1", "pDC")
Mk <- c("Megakaryocyte", "Megakaryocyte progenitor", "Platelet")

progenitor_cells <- c("CLP", "CMP", "GMP", "MEP", "MPP", "CDP", "HSC")

stroma <- c("Fibroblast", "Mesenchymal stromal cell", "Osteoclast")
```
# plot cell frequency plots
```{r plot cell frequency}
simplified_ct_list <- rna.data@meta.data %>% distinct(simplified_ct, ct)
simplified_ct_plot_list <- list()
Idents(rna.data) <- rna.data$simplified_ct
# plot with for loop
for (i in seq_along(simplified_ct_list$simplified_ct %>% unique())) {
  cell_type <- unique(simplified_ct_list$simplified_ct)[i]
  print(cell_type)
  ct_list <- simplified_ct_list %>%
    filter(simplified_ct == cell_type) %>%
    pull(ct)
  dittoFreqPlot(rna.data, "ct", sample.by = "sample", # not sample id
                group.by = "Predicted_sex", data.out = TRUE,
                vars.use = ct_list,

                main = paste0(cell_type, " frequency")) -> freq.data
  freq.data[[2]] %>%
    filter(label.count.total.per.facet > 10) %>% # remove facet points with less than 10 cells
    ggplot(aes(x = grouping, y = percent)) +
    geom_boxplot() +
    geom_point() +
    facet_wrap(~label, scales = "free") +
    labs(title = paste0(cell_type, " frequency under 45")) -> plot

  assign(paste0(gsub(" ", "_", cell_type), "_plot"), plot)
  simplified_ct_plot_list <- c(simplified_ct_plot_list, mget(paste0(gsub(" ", "_", cell_type), "_plot")))
}
# plot frequency of all cell types by sample faceted by report
rna.data %>%
  dittoBarPlot(var = "ct", group.by = "sample", split.by = "projectId",
               split.adjust = list(scales = "free"), # removes empty groups
               main = "Cell type by individual and project"
  ) -> ct_individual_plot
ct_individual_plot

dittoBarPlot(rna.data, var="ct", group.by="sample", scale = "count",
             split.by = "projectId", split.adjust = list(scales = "free")) -> progenitor_counts
progenitor_counts
```
For specific plots:
```{r}
cell_type_list <- c("CLP", "MPP", "CMP", "GMP", "MEP", "CDP", "HSC")
ct_plot_list <- list()
for (i in seq_along(cell_type_list)) {
  cell_type <- cell_type_list[i]
  print(cell_type)
  dittoFreqPlot(rna.data, "ct", sample.by = "sample", # not sample id
              group.by = "Predicted_sex", data.out = TRUE, vars.use = cell_type,
                main = paste0(cell_type, " frequency")) -> freq.data
  freq.data[[2]] %>%
    filter(label.count.total.per.facet > 10) %>% # remove facet points with less than 10 cells
    ggplot(aes(x = grouping, y = percent, fill=grouping)) +
    geom_violin() +
    geom_point() +

    facet_wrap(~label, scales = "free") +
          # change name of x axis
          xlab("Sex") +
            # change name of y axis
            ylab("Frequency overall") +

          theme_bw() +
          # big text
            theme(text = element_text(size = 20)) +

    labs(title = paste(cell_type, "frequency")) -> plot
  assign(paste0(gsub(" ", "_", cell_type), "_plot"), plot)
  ct_plot_list <- c(ct_plot_list, mget(paste0(gsub(" ", "_", cell_type), "_plot")))
}
```

```{r}
pdf("individual_celltype_distribution.pdf", width = 10, height = 10)
ct_individual_plot
simplified_ct_plot_list
ct_plot_list
dev.off()
ct_individual_plot
simplified_ct_plot_list
ct_plot_list
```

# speckle
```{r speckle}
propeller(clusters = rna.data$ct, sample = rna.data$sample, group = rna.data$Predicted_sex) -> prop.out
prop.out
propeller(clusters = rna.data$ct, sample = rna.data$sample, group = rna.data$Predicted_sex,
          transform = "asin"
) -> propeller_asin.out
propeller_asin.out
```
make dot plot
```{r}
# add log2 ratio
prop.out$PropRatioLog2 <- log2(prop.out$PropRatio)
prop.out %>%
  # filter(P.Value < 0.05) %>%
  # sort by PropRatioLog2 for ggplot
  arrange(desc(PropRatioLog2)) %>%
  mutate(BaselineProp.clusters = factor(BaselineProp.clusters, levels = BaselineProp.clusters)) %>%
  # make a dot plot with ggplot geom_point
  ggplot(aes(x = PropRatioLog2, y = BaselineProp.clusters)) +
  scale_size(trans = "reverse", breaks = c(0.05, 0.1, 0.2, 0.5, 0.75)) +

geom_point(aes(size = P.Value, color = I(ifelse(P.Value < 0.05, "green",
                                                ifelse(P.Value < 0.1, "yellow", "red"))))) +  theme_bw() +
  # theme(legend.position = "none") +
  labs(title = "Cell type distribution") -> dotplot
dotplot
# same with propeller_asin.out
propeller_asin.out$PropRatioLog2 <- log2(propeller_asin.out$PropRatio)

propeller_asin.out %>%
  # filter(P.Value < 0.05) %>%
  # sort by PropRatioLog2 for ggplot
  arrange(desc(PropRatioLog2)) %>%
  mutate(BaselineProp.clusters = factor(BaselineProp.clusters, levels = BaselineProp.clusters)) %>%
  # make a dot plot with ggplot geom_point
  ggplot(aes(x = PropRatioLog2, y = BaselineProp.clusters)) +
  scale_size(trans = "reverse", breaks = c(0.05, 0.1, 0.2, 0.5, 0.75)) +

geom_point(aes(size = P.Value, color = I(ifelse(P.Value < 0.05, "green",
                                                ifelse(P.Value < 0.1, "yellow", rgb(1,0,0,0.25)))))) +

  theme_bw() +

  # theme(legend.position = "none") +
  labs(title = "Cell type distribution") -> dotplot_asin
dotplot_asin
```

Plot freq m vs f
```{r}
plot(prop.out$PropMean.M, prop.out$PropMean.F)
# with ggplot
prop.out %>%
  ggplot(aes(x = PropMean.M, y = PropMean.F)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  # add labels
  geom_text(aes(label = BaselineProp.clusters), hjust = 0, vjust = 0) +
  labs(title = "Cell type distribution") -> freq_plot
freq_plot
# log2 ratio
plot(log(prop.out$PropMean.M), log(prop.out$PropMean.F))
lines(c(0, 0), c(1, 1))
# with ggplot
prop.out %>%
  ggplot(aes(x = log(PropMean.M), y = log(PropMean.F))) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  # add labels
  geom_text_repel(aes(label = BaselineProp.clusters), hjust = 0, vjust = 0) +
  labs(title = "Cell type distribution") -> freq_plot
freq_plot
# only show progenitors
prop.out %>%
  filter(BaselineProp.clusters %in% progenitor_cells | BaselineProp.clusters %in% dendritic_cells) %>%
  ggplot(aes(x = log(PropMean.M), y = log(PropMean.F))) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  # add labels
  geom_text_repel(aes(label = BaselineProp.clusters), hjust = 0, vjust = 0) +
  labs(title = "Cell type distribution") -> freq_plot
freq_plot
```

# now with simplified_ct
Not for just progenitors
```{r, eval = FALSE}
propeller(clusters = rna.data$simplified_ct, sample = rna.data$sample, group = rna.data$Predicted_sex) -> sct_prop.out
sct_prop.out
sct_prop.out$PropRatioLog2 <- log2(sct_prop.out$PropRatio)
sct_prop.out %>%
  # filter(P.Value < 0.05) %>%
  # sort by PropRatioLog2 for ggplot
  arrange(desc(PropRatioLog2)) %>%
  mutate(BaselineProp.clusters = factor(BaselineProp.clusters, levels = BaselineProp.clusters)) %>%
  # make a dot plot with ggplot geom_point
  ggplot(aes(x = PropRatioLog2, y = BaselineProp.clusters)) +
  geom_point(aes(size = P.Value, color = I(ifelse(P.Value < 0.05, "green",
                                                ifelse(P.Value < 0.1, "yellow", "red"))))) +
  scale_size(trans = "reverse", breaks = c(0.05, 0.1, 0.2, 0.5, 0.75)) +
  theme_bw() +
  # theme(legend.position = "none") +
  labs(title = "Cell type distribution") -> sct_dotplot
sct_dotplot
```

# disco
```{r}
bm <- readRDS("../data/DISCO/bmv2_fixed.rds")
# not NA in gender
bm <- subset(bm, subset= gender != "")
propeller(clusters =bm$ct, sample = bm$sample, group = bm$gender) -> bm_prop.out
bm_prop.out
propeller(clusters = bm$ct, sample = bm$sample, group = bm$gender,
          transform = "asin"
) -> bm_propeller_asin.out
bm_propeller_asin.out
# add log2 ratio
bm_prop.out$PropRatioLog2 <- log2(bm_prop.out$PropRatio)
bm_propeller_asin.out$PropRatioLog2 <- log2(bm_propeller_asin.out$PropRatio)
```


```{r}
bm_prop.out %>%
  # filter(P.Value < 0.05) %>%
  # sort by PropRatioLog2 for ggplot
  arrange(desc(PropRatioLog2)) %>%
  mutate(BaselineProp.clusters = factor(BaselineProp.clusters, levels = BaselineProp.clusters)) %>%
  # make a dot plot with ggplot geom_point
  ggplot(aes(x = PropRatioLog2, y = BaselineProp.clusters)) +
  scale_size(trans = "reverse", breaks = c(0.05, 0.1)) +

  geom_point(aes(size = P.Value, color = I(ifelse(P.Value < 0.05, "green",
                                                ifelse(P.Value < 0.1, "yellow", "red"))))) +
  # theme(legend.position = "none") +
  labs(title = "Cell type distribution") -> bm_dotplot
bm_dotplot
```


# issues

when taking only the progenitors from the start, too few cells are left for the analysis. This creates issues during the scaling function.
Maybe we should focus on other datasets.
```{r}
rna.data@meta.data %>%
filter(simplified_ct == "Progenitor cells") %>%
group_by(sample, projectId) %>%
summarise(n=n()) -> counts
counts
# change NA in projectId to HCA
counts$projectId[is.na(counts$projectId)] <- "HCA"
dittoBarPlot(subset(rna.data, subset = simplified_ct == "Progenitor cells"), var="ct", group.by="sample")
dittoBarPlot(subset(rna.data, subset = simplified_ct == "Progenitor cells"), var="ct", group.by="sample", scale = "count")
# stacked counts
counts %>%
  ggplot(aes(x = sample, y = n)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  # facet by projectId
    facet_wrap(~projectId, scales="free") +

  labs(title = "Progenitor cells counts") -> progenitor_counts
progenitor_counts
```

Idem but for disco
```{r}
# with bm
bm@meta.data %>%
  filter(simplified_ct == "Progenitor cells") %>%
  group_by(sample, cell_sorting) %>%
  summarise(n=n()) -> bm_counts
bm_counts
bm_counts %>%
  ggplot(aes(x = sample, y = n)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  # add text with cell sorting
    geom_text(aes(label = cell_sorting), hjust = 0, vjust = 0) +
  labs(title = "Progenitor cells counts") -> bm_progenitor_counts
bm_progenitor_counts
dittoBarPlot(subset(bm, subset = simplified_ct == "Progenitor cells"), var="ct", group.by="sample")
```
